---
title: "Aluminum Data"
author: "Jacob Smilg"
date: 2023-02-15
output:
  github_document:
    toc: true
prerequisites:
  - e-data03-pivot-basics
---

*Purpose*: When designing structures such as bridges, boats, and planes, the design team needs data about *material properties*.
Often when we engineers first learn about material properties through coursework, we talk about abstract ideas and look up values in tables without ever looking at the data that gave rise to published properties.
In this challenge you'll study an aluminum alloy dataset: Studying these data will give you a better sense of the challenges underlying published material values.

In this challenge, you will load a real dataset, wrangle it into tidy form, and perform EDA to learn more about the data.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**.
The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|--------------------------|-----------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge.
See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a number of Aluminum plates to determine their [elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and [Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio).
These are key quantities used in the design of structural members, such as aircraft skin under [buckling loads](https://en.wikipedia.org/wiki/Buckling).
These scientists tested plats of various thicknesses, and at different angles with respect to the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking)) direction.

# Loading and Wrangle

<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data form many sources.
The `read_csv()` function will help us load the data for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang
```

Note that these data are not tidy!
The data in this form are convenient for reporting in a table, but are not ideal for analysis.

### **q1** Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, nu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special value for `names_to`.

```{r q1-task}
## TASK: Tidy `df_stang`
df_stang_long <-
  df_stang %>% 
  pivot_longer(
    names_to = c(".value", "angle"),
    names_sep = "_",
    starts_with("E") | starts_with("nu")
  ) %>% 
  filter(E >= 0) %>% 
  mutate(angle = as.integer(angle))

df_stang_long

```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "nu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(26, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

# EDA

<!-- -------------------------------------------------- -->

## Initial checks

<!-- ------------------------- -->

### **q2** Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own *specific* question that you'd like to answer about the data---you'll answer it below in q3.

```{r q2-task}

df_stang_long %>% 
  group_by(alloy) %>% 
  summarise(
    nu_sd = sd(nu),
    nu_mean = mean(nu),
    nu_median = median(nu),
    nu_cov = nu_sd / nu_mean,
    
    E_sd = sd(E),
    E_mean = mean(E),
    E_median = median(E),
    E_cov = E_sd / E_mean,
  )

df_stang_long %>% 
  group_by(thick) %>% 
  summarise(
    nu_sd = sd(nu),
    nu_mean = mean(nu),
    nu_median = median(nu),
    nu_cov = nu_sd / nu_mean,
    
    E_sd = sd(E),
    E_mean = mean(E),
    E_median = median(E),
    E_cov = E_sd / E_mean,
  )

df_stang_long %>% 
  group_by(angle) %>% 
  summarise(
    nu_sd = sd(nu),
    nu_mean = mean(nu),
    nu_median = median(nu),
    nu_cov = nu_sd / nu_mean,
    
    E_sd = sd(E),
    E_mean = mean(E),
    E_median = median(E),
    E_cov = E_sd / E_mean,
  )

```

**Observations**:

-   Is there "one true value" for the material properties of Aluminum?
    -   Maybe, but it's not super clear. In general, `nu` and `E` seem to vary a little bit with `angle` or `thick` for the three smaller thicknesses, and more significantly for the 0.081 thickness. I think this probably indicates that there isn't "one true value" - if there were, `nu` and `E` shouldn't vary as much.
-   How many aluminum alloys are in this dataset? How do you know?
    -   There's only one unique value in the `alloy` column, so there's only one alloy in the dataset.
-   What angles were tested?
    -   0, 45, and 90 degrees.
-   What thicknesses were tested?
    -   0.022, 0.032, 0.064, and 0.081 inch.
-   Do `E` and `nu` vary with angle?
    -   Not by much --- the medians by angle are almost the same (0 and 90 deg have the same median, and 45 deg median is only 50 kips/in\^2 larger.

## Visualize

<!-- ------------------------- -->

### **q3** Create a visualization to investigate your question from q2 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
df_stang_long %>% 
  ggplot(aes(x = E, y = nu, color = as_factor(angle))) + 
  geom_point(size = 3) +
  geom_smooth(method=lm, se=FALSE)
```

**Observations**:

-   There doesn't appear to be any strong relationship between `angle` and `E` or `nu`. The best fit lines are all very close together - the 45 degree line does appear to correspond to higher `nu` values, but there's not enough data to draw any solid conclusions. A pattern may or may not appear with more data.

### **q4** Consider the following statement:

"A material's property (or material property) is an intensive property of some material, i.e. a physical property that does not depend on the amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a tested plate.
Does the following graph support or contradict the claim that "elasticity `E` is an intensive material property." Why or why not?
Is this evidence *conclusive* one way or another?
Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%
  ggplot(aes(nu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

-   Does this graph support or contradict the claim above?
    -   By itself, the graph appears to contradict the claim that `E` is an intensive property.
        The data for the 0.081" thickness is concentrated in the bottom left corner, which alone would seem to contradict the claim. However, there is little context provided for the graph or data --- it's possible that the range in which `nu` and `E` vary is acceptably small for many applications to consider `E` to be intensive.

    -   Also, the paper that the data is from mentions that the values for 0.081" thickness were collected with different measurement devices, which could have introduced some error that caused the `E` and `nu` values to skew lower. Considering those are the only values that are clearly separate from the others, I'm suspicious of the quality of the data.

    -   Basically, I don't think you can make a strong case for or against the claim based on the above graph alone.

# References

<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural alloys for large strains" (1946) Journal of Research of the National Bureau of Standards, (pdf link)[<https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf>]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26, (link)[<https://en.wikipedia.org/wiki/List_of_materials_properties>]
